---
title: 'Comparing the diferrence between two means: Student''s t-test'
author: "Dr. Peter McKenna"
date: "22 August 2018"
output: ioslides_presentation
---


```{r setup, include=F}
knitr::opts_chunk$set(echo = F)

setwd("~/User files/Peter/Heriot-Watt University/R Statistics/HRI statistics course/HRI stats course/interaction-design")

```


## **Introduction**

> - In this class you are going to learn how to perform a t-test using R Studio
> - A t-test is a statistical test that tests whether the difference between to sample means is significantly differnt
> - This test is often used in human-robot and human-computer studies
> - You may need to use this test on your own data for your 4th year project



## **Loading the necessary packages**

- First, open R Studio
- To begin, we must load a set of packages that will allow us to wrangle (i.e. tidy, sort), explore, and analyse the data
- To do this we need to load the `dplyr`, `tidyr`, `readr`, and `ggplot2` packages with the following syntax

```{r loading packages, echo = T, warning = F, message = F}

library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)

```

- do not be concerned by the text that is generated in the console; R packages are stamped to certain versions of R Studio, but work fine once loaded



## **Main verbs of R's Tidyverse**

- Here is a list of some of the functions we will be using today.
- The code is ready for you to use, so use these as a reference point in the proceeding slides

>- `filter`: extract rows
>- `select`: extract columns 
>- `gather/spread`: gather columns into rows/spread rows into columns
>- `mutate`: compute and append new/existing columns
>- `summarise`: summarise data based on stated criteria
>- see online text "R for Data Science" for more info 



## Retrieving the data

- the data and code you need for today can be found here

https://github.com/petermckenna/interaction-design

- on this page click **Clone or download** --> **Download ZIP**

- Save the Zip folder in your user directory, e.g. C/Users/xm42/University/Interaction Design. You may need to copy it from your *Downloads* folder

- Save the folder where you keep your other **Interaction Design** materials



## Setting up a working directory pt 1

- to set up the directory locate the Interaction Design Zip folder by...

>- Going to R Studio and clicking the **Files** tab on the right panel

>- Click the 3 dots **...** below the refresh symbol

>- Locate the Zip folder and click **OK**

>- Click the **More** option to the left of the **...** button and select **Set as Working Directory**



## Setting up a working directory pt 2

- Alternatively, enter the directory manually using the command `setwd` and quoting the file path, e.g.

`setwd("C/Users/xm42/University/Interaction Design")`


## **Reading in the data**

- We then read in the data (a .csv file) useing the following command

```{r reading in the data, echo=TRUE, warning = F, message = F}

t_test_data <- read_csv("t-test-data.csv")

```

- this syntax creates an object called `t_test_data` from an existing .csv data file


## **Viewing the data**

- Let's have a look at the data
- Do do this you can clik on the `t_test_data` object created in the global environment (panel in top right)
- or we can use the `head` command to examine first 6 rows of the data 


---

```{r viewing the data, echo = T}

head(t_test_data)

```


## **Description of the data** {.smaller}

> - The data examines 60 participants Likert rating and response time of animated v non-animated robot
> - Each row represent a different participant responses, so there are 60 rows 
> - `id` = participant number
> - `Likert_rating` = participants evaluation of the inteaction from 1 (Not very friendly) to 5 (very friendly)
> - `animacy` = condition with two levels (animated, non-animated)
  > - in the **animated** conditon the robot tracks participant's heads and gestures in communication
  > - in the **non-animated** condition the robot is static
> - `response_time` vector = the time (seconds) it took for participants to indicate how friendly they thought the robot was on the Likert scale
> - the experiment (user study) is between-subjects; 
> - i.e., one set of paticipants interact with the animated robot, and a separate group interact with the non-animated robot


## **Generating descriptve statistics** 

- To generate a summary the data we use the `summary` command

```{r summarising data, echo = T}  

summary(t_test_data)

```

- as you can see, `summary` has generated the min, max, median, and mean for our outcome vectors `likert_num` and `response_time` 



## **Generating descriptve statistics | Between group summaries: part 1**

> - We are interested in differences between the participants who interacted with the **animated** and the **non_animated** robot
> - So, we generate summaries the outcome variables (`likert_num`, `response_time`) by group (**animated**, **non_animated**) 
> - The following syntax generated the mean and standard deviation for each combination of group and outcome variable 


## **Generating descriptve statistics | Between group summaries: part 2**

```{r summarising by group, echo = T}

t_test_data %>%
  group_by(animacy) %>%                     
  summarise(avg_rating = mean(likert_num),  
            sd_rating = sd(likert_num),
            avg_rt = mean(response_time),
            sd_rt = sd(response_time))

```

## **Generating descriptve statistics | Between group summaries: part 3**

- As you can see, particpants in the **animated** robot group rated the robot as more friednly (M = 4.13, SD = 0.86) compared to participants in the **non-animated** robot group (M = 2.77, 1.01).  
- Also, particpants in the **animated** robot group took longer to respond to the Likert question on Friendliness (M = 5.06, SD = 1.06) compared to participants in the **non-animated** robot group (M = 3.46, 1.03). 


## **Checking for normality**

>- Deciding what statistical test to use depends on the distribution of the data
>- Here is the basic rule of thumb: a bell-shaped = normally distributed (known as *parametic*)
>- non-bell shaped (skewed) data is non-parametric.
>- For continuous data (e.g. interval, ratio scale) we plot a hisogram to check shape of distribution

## **A parametric distribution**

- This is what is known as a **Gaussian** (or parametric) distribution
- You can see why it is referred to as bell-shaped

```{r plot of normal distribution, echo = F}

x <- seq(5,15,length=1000)
y <- dnorm(x,mean=10, sd=3)
plot(x,y, type="l", lwd=1)

```

## **A non-parametric distribution**

- And this is what a non-paramettric distribution looks like
- Notice that the area under the line is not symmetrical: an indication that the data is non-parametric

```{r plot of non-normal distribution, echo = F, warning=F}

n <- 2000
k <- seq(0, n, by = 20)
plot (k, dbinom(k, n, pi/10, log = TRUE), type = "l", ylab = "log density")


```


## **Check distribution of resp_time data part 1**

- Distribution of non-animated robot data

```{r normality check pt1, echo = T, eval = F}

t_test_data %>%
  filter(animacy == "non_animated") %>%
  ggplot(.) +
  geom_histogram(mapping = aes(x = response_time),
                 bins = 20)


```

----

```{r normality check pt2, echo = F}

t_test_data %>%
  filter(animacy == "non_animated") %>%
  ggplot(.) +
  geom_histogram(mapping = aes(x = response_time),
                 bins = 20)


```


## **Check distribution of resp_time data part 2**

- Distribution of animated robot data

```{r normality check pt3, echo = T, eval = F, warning = F, message = F}

t_test_data %>%
  filter(animacy == "animated") %>%
  ggplot(.) +
  geom_histogram(mapping = aes(x = response_time),
                 bins = 20)

```

----

```{r normality check pt4, echo = F, warning = F, message = F}

t_test_data %>%
  filter(animacy == "animated") %>%
  ggplot(.) +
  geom_histogram(mapping = aes(x = response_time),
                 bins = 20)

```

## **Checking for group (animated, non_animated) differences**


- A usful way to check the differnce in scores between groups is to generate a boxplot
- We use the `ggpot2` package to generate graphics
- For a boxplot, the syntax is geom_boxplot
- First we'll look at `response_time` then `likert_num`


## **Response Time(sec) between groups**

```{r checking group differences in rt with boxplot, echo = T}

t_test_data %>%
  ggplot(.) +
  geom_boxplot(mapping = aes(x = animacy, y = response_time, fill = animacy)) + 
  theme(legend.position = "none")

```

- in a boxplot the bold lines in the box show the median
- There is a big difference between the medians of the two groups and their response time (sec)

## **Rating of robot's friendliness between groups**

```{r checking group differences in likert with boxplot, echo = T}

# first, make an object (here I've called it 't') that contains the means and sd of likert_num

t <- t_test_data %>%
  group_by(animacy) %>%                     
  summarise(avg_rating = mean(likert_num),  
            sd_rating = sd(likert_num))

# then pass to ggplot2

t %>%
  ggplot(., aes(animacy, avg_rating, fill = animacy)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin = avg_rating - sd_rating, ymax = avg_rating + sd_rating), width = .2)


```

## **Performing a t-test | pt 1 likert data**

>- Now we can check if our groups differ significantly in terms of how friednly they rated the robot (**likert_num**) and their response time (**response_time**)
>- We'll start with the Likert data
>- The `t.test` will calculate this difference
>- To pass data to `t.test` we need to remove `id` and `response_time` vectors to perform grouping, then add `id` again at the end

----

```{r t-test likert data, echo T }

t_test <- t_test_data %>%
  select(-id, -response_time) %>%  
  group_by_at(vars(animacy)) %>%
  mutate(id=1:n()) %>% 
  ungroup() %>% 
  spread(key = animacy, value = likert_num) 

```

----

- This new object `t_test` can no be passed to the `t.test` function
- We use the `$` to specify a vector from the object, in this case, our groups

```{r t-test likert analysis, echo = T}

t.test(t_test$non_animated, t_test$animated)

```
>- Note: if want paired samples t-test add `paired=FALSE`, e.g. t.test(t_test$non_animated, t_test$animated, paired = FALSE) 


## **Interpreting the output**

- Your interested in the 3rd line of output ( the one beginning with *t*)
- t = -5.6539, df = 56.632, p-value = 5.357e-07
- So, p < 0.001.
- Convention holds that this is statisically significant
- So you can conclude that particpants in the **animated** robot group rated the robot as significantly friendlier (M = 4.13, SD = 0.86) compared to participants in the **non-animated** robot group (M = 2.77, 1.01), *t* = -5.6539, df = 56.632, *p* < 0.001. 


## **Performing a t-test | pt 2 response time data**

- This time a new object called `t_test1` is created for the `response_time` data t-test

```{r t-test response time, echo = T}

t_test1 <- t_test_data %>%
  select(-id, -likert_num) %>%  
  group_by_at(vars(animacy)) %>%
  mutate(id=1:n()) %>% 
  ungroup() %>% 
  spread(key = animacy, value = response_time) 


```

----

- Then we pass this data to the t.test function and interpret the findings in the same way

```{r}

t.test(t_test1$non_animated, t_test1$animated)

```

