---
title: 'Comparing the diferrence between two means: Student''s t-test'
author: "Dr. Peter McKenna"
date: "22 August 2018"
output: ioslides_presentation
---

<style>
pre {
  font-size: 16px;
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(echo = F)



```


## **Introduction**

> - In this class you are going to learn how to perform a t-test using R Studio
> - A t-test is a statistical test that tests whether the difference between to sample means is significantly differnt
> - This test is often used in human-robot and human-computer studies
> - You may need to use this test on your own data for your 4th year project

## **Loading the necessary packages**

- To begin, we must first load a set of packages that will allow us to wrangle (i.e. tidy, sort), explore, and analyse the data
- To do this we need to load the `dplyr`, `tidyr`, `readr`, and `ggplot2` packages with the following syntax

```{r loading packages, echo = T, warning = F, message = F}

library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)

```

- do not be concerned by the text that is generated in the console; R packages are stamped to certain versions of R Studio, but work fine once loaded

## **Main verbs of R's Tidyverse**

- `filter`: extract rows
- `select`: extract columns 
- `gather/spread`: gather columns into rows/spread rows into columns
- `mutate`: compute and append new/existing columns
- `summarise`: summarise data based on stated criteria
- see online text "R for Data Science" for more info 


## **Reading in the data**

- We then read in the data (a .csv file) useing the following command

```{r reading in the data, echo=TRUE, warning = F, message = F}

t_test_data <- read_csv("t-test-data.csv")

```

- this syntax creates an object called `t_test_data` from an existing .csv data file


## **Viewing the data**

- Let's have a look at the data
- Do do this you can clik on the `t_test_data` object created in the global environment (panel in top right)
- or we can use the `head` command to examine first 6 rows of the data 

```{r viewing the data, echo = T}

head(t_test_data)

```

## **Description of the data** {.smaller}

> - The data examines 60 participants Likert rating and response time of animated v non-animated robot
> - Each row represent a different participant responses, so there are 60 rows 
> - *id* vector = participant number
> - *Likert_rating* vector = participants evaluation of the inteaction from 1 (Not very friendly) to 5 (very friendly)
> - *animacy* vector = condition with two levels (animated, non-animated)
> - in the animated conditon the robot tracks participant's heads and gestures in communication
> - in the non-animated condition the robot is static
> - *response_time* vector = the time (seconds) it took for participants to indicate how friendly they thought the robot was on the Likert scale
> - the experiment (user study) is between-subjects; 
> - i.e., one set of paticipants interact with the animated robot, and a separate group interact with the non-animated robot


## Generating descriptve statistics 

- to generate a general summary the data we use the `summary` command

```{r summarising data, echo = T}  

summary(t_test_data)

```

- as you can see, summary has generated the min, max, median, IQR and mean for our outcome vectors `likert_num` and `response_time` 



## Generating descriptve statistics | Between group summaries: part 1 

> - We are interested in differences between the participants who interacted with the `animated` and the `non_animated` robot
> - So, we generate summaries the outcome variables (`likert_num`, `response_time`) by group (`animated`, `non_animated`) 
> - The following syntax generated the mean and standard deviation for each combination of group and outcome variable 


## Generating descriptve statistics | Between group summaries: part 2

```{r summarising by group, echo = T}

t_test_data %>%
  group_by(animacy) %>%                     
  summarise(avg_rating = mean(likert_num),  
            sd_rating = sd(likert_num),
            avg_rt = mean(response_time),
            sd_rt = sd(response_time))

```

## Generating descriptve statistics | Between group summaries: part 3

- As you can see, particpants in the `animated` robot group rated the robot as more friednly (M = 4.13, SD = 0.86) compared to participants in the `non-animated` robot group (M = 2.77, 1.01).  
- Also, particpants in the `animated` robot group took longer to respond to the Likert question on Friendliness (M = 5.06, SD = 1.06) 
- compared to participants in the `non-animated` robot group (M = 3.46, 1.03). 


## Checking for normality

- Deciding what statistical test to use depends on the distribution of the data
- Here is the basic rule of thumb: a bell-shaped = normally distributed (known as *parametic*)
- non-bell shaped (skewed) data is non-parametric.
- For continuous data (e.g. interval, ratio scale) we plot a hisogram to
- check shape of distribution



